### 2023-01-10 appendix.Rmd
including Tim's comments

```{r Figure2, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by *M*."}
### For the appendix -- other splits for the main results as plot (N is most important)
# Plot the results, by M
ldf %>% 
  group_by(method, Objective, M) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>% 
  ggplot(aes(x = M, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 2:5)
```

```{r Figure3, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by *SD*."}
# Plot the results, by SD
ldf %>% 
  group_by(method, Objective, SD) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>% 
  ggplot(aes(x = SD, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 1:3)
```

```{r Figure4, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by *K*."}
# Plot the results, by K
ldf %>% 
  group_by(method, Objective, K) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>% 
  ggplot(aes(x = K, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 2:4)
```

```{r Figure5, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by *r*."}
# Plot the results, by r
ldf %>% 
  group_by(method, Objective, r) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>% 
  ggplot(aes(x = r, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 0:5 / 10)
```

>>>>

```{r Figure2, fig.width = 8, fig.height = 10, fig.cap = "Results of the simulation study: Depicts the performance of the anticlustering methods with regard to minimizing discrepancy in means, variances, skewness, kurtosis and correlations (split by the number of features, averaged across the remaining variables that varied in the simulation)."}
### For the appendix -- other splits for the main results as plot (N is most important)
# Plot the results, by M (number of features)
ldf %>% 
  group_by(method, Objective, M) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = paste0(
        "Delta[italic(",
        c("M", "SD", "Skew", "Kurtosis", "Cor"),
        ")]"
      )
    ) 
  ) %>% 
  ggplot(aes(x = M, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free", labeller = label_parsed) + 
  ylab("Mean discrepancy") +
  xlab("Number of features") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 2:5)
```

```{r Figure3, fig.width = 8, fig.height = 10, fig.cap = "Results of the simulation study: Depicts the performance of the anticlustering methods with regard to minimizing discrepancy in means, variances, skewness, kurtosis and correlations (split by *SD*, averaged across the remaining variables that varied in the simulation)."}
# Plot the results, by SD
ldf %>% 
  group_by(method, Objective, SD) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = paste0(
        "Delta[italic(",
        c("M", "SD", "Skew", "Kurtosis", "Cor"),
        ")]"
      )
    ) 
  ) %>% 
  ggplot(aes(x = SD, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free", labeller = label_parsed) + 
  ylab("Mean discrepancy")+
  xlab(expression(italic(SD))) +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 1:3)
```

```{r Figure4, fig.width = 8, fig.height = 10, fig.cap = "Results of the simulation study: Depicts the performance of the anticlustering methods with regard to minimizing discrepancy in means, variances, skewness, kurtosis and correlations (split by *K*, averaged across the remaining variables that varied in the simulation)."}
# Plot the results, by K
ldf %>% 
  group_by(method, Objective, K) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = paste0(
        "Delta[italic(",
        c("M", "SD", "Skew", "Kurtosis", "Cor"),
        ")]"
      )
    ) 
  ) %>% 
  ggplot(aes(x = K, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free", labeller = label_parsed) + 
  ylab("Mean discrepancy")+
  xlab(expression(italic(K))) +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 2:4)
```

```{r Figure5, fig.width = 8, fig.height = 10, fig.cap = "Results of the simulation study: Depicts the performance of the anticlustering methods with regard to minimizing discrepancy in means, variances, skewness, kurtosis and correlations (split by *r*, averaged across the remaining variables that varied in the simulation)."}
# Plot the results, by r
ldf %>% 
  group_by(method, Objective, r) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = paste0(
        "Delta[italic(",
        c("M", "SD", "Skew", "Kurtosis", "Cor"),
        ")]"
      )
    ) 
  ) %>% 
  ggplot(aes(x = r, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free", labeller = label_parsed) + 
  ylab("Mean discrepancy")+
  xlab(expression(italic(r))) +
  theme_bw(base_size = 16) + 
  scale_x_continuous(breaks = 0:5 / 10)
```

### 2023-01-10 paper.Rmd
including Tim's comments

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximizing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), while the classical k-means criterion only reflects the degree to which groups differ in their means. It is shown that the k-plus criterion can be reduced to an augmentation of the input data while the original k-means objective remains unchanged. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances usually does not compromise similarity with regard to means; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Example are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

# Compute features for k-means-extended criterion
squared_from_mean <- function(data) {
  apply(data, 2, function(x) (x - mean(x))^2)
}

@@

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity, and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of dissimilarities between elements within the same group [@brusco2019; @gallego2013]. By considering pairwise dissimilarities from between elements in the same group, the diversity reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion based on the information in a dissimilarity matrix is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases within-group heterogeneity by ensuring that any two elements in the same group as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications striving for within-group heterogeneity should incorporate both dispersion and diversity, and they presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals oftentimes coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. In the context of anticlustering, one approach for handling attribute data is to compute an appropriate dissimilarity measure such as the pairwise Euclidean or squared Euclidean distance across the set of attributes [@brusco2019]. Subsequently, a distance-based anticlustering criterion such as the diversity can be optimized to obtain a partitioning of the stimuli. A different anticlusting approach uses clustering objectives that can directly be computed on the attribute values. K-means is probably the best-known objective of this kind. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, the negligence of distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

If the $\mathit{SSE}$ is maximal---which is the goal of k-means *anticlustering*---, the cluster centroids are as close as possible to the overall centroid $\overline{\mathbf{x}}$ and therefore to each other [@spath1986]. Thus, k-means anticlustering directly optimizes the similarity of the mean attribute values across clusters.

@@

To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. This simplification leads to the *unweighted sum approach* for k-plus anticlustering: When defining $z_{i.} = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, x_{iP})$, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

@@

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged. Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardize before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

```{r, fig.cap = "The relationship between the objectives $SSE$ and $SSE_{Var}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the triangular minimizes the difference in variances; the partition represented by the square minimizes the difference in means; the partition represented by the diamond maximizes $SSE_{kplus}$ and satisfies both criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

@@

The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means; (ii) diversity anticlustering, maximizing the sum of Euclidean distances within groups; (iii) a simple random allocation of elements to groups. All anticlustering methods used a $z$-standardization of the data sets as a preprocessing step.

@@

apa_table(
  tab,
  caption = "Results of the simulation study",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average between-group discrepancy with regard to means, standard deviations, skewness, kurtosis, and correlations. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

```{r Figure1, fig.width = 8, fig.height = 10, fig.cap = "Between-group discrepancies for all evaluation criteria split by $N$."}

@@

ldf %>% 
  group_by(method, Objective, N) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = c("M", "SD", "Skew", "Kurtosis", "Correlation"))
  ) %>%
  ggplot(aes(x = N, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free") + 
  ylab("Mean discrepancy") +
  theme_bw(base_size = 16) + 
  scale_x_continuous(
    breaks = axis_ticks, 
    labels = axis_labels
  )

@@

I used k-plus anticlustering to divide the data set of 900 images into 9 groups of 100 images each. The task was accomplished using the following R code:

@@

kplus_anticlustering(
  features,
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  covariances = FALSE,
  moments = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```

The code runs in about 10 seconds on a contemporary personal computer; the run time may vary depending on how many iterations are needed to find the local maximum. The variable `features` is a data table containing the image data: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. The accompanying OSF repository contains code to reproduce the example (**TODO**).

The `kplus_anticlustering()` function has four boolean arguments to specify whether variance, skewness, kurtosis and covariances should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness, kurtosis and convariance have to be spefically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features are standardized before the optimization starts; in particular, this option enforces that all k-plus criteria receive the same weight during the optimization process. This behaviour can be adjusted using the boolean argument `standardize`. The documentation of the `anticlust` package also provides detailed information on how to use the function `kplus_anticlustering()`.

@@

Table 4 illustrates the results that are obtained when using standard k-plus anticlustering to divide the stimulus set into 6 groups of size `r nrow(schaper2019) / 6` each. Even though the means and standard deviations are no longer perfectly matched between groups, arguably they are similar enough for any practical purposes.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: mean, variances, skewness, kurtosis, higher order moments, and covariances. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, however, the k-plus extension does not change the original k-means criterion because additional criteria can be incorporated by augmenting the input data.

A simulation study and examples on real norming data showed that the k-plus criterion is well suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be aggravated in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Unlike in statistical analysis, there is no problem with "just trying out" different angles when it comes to anticlustering. Moreover, using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.

>>>>

    Anticlustering refers to the process of partitioning elements into disjoint groups with the goal of obtaining high between-group similarity and high within-group heterogeneity. Anticlustering thereby reverses the logic of its better known twin---cluster analysis---and is usually approached by maximizing instead of minimizing a clustering objective function. Introducing k-plus, this paper presents an extension of the k-means clustering objective to maximize between-group similarity in anticlustering applications. The k-plus criterion represents between-group similarity as discrepancy in distribution moments (means, variance, higher order moments, and covariances), while the classical k-means criterion only reflects the degree to which groups differ in their means. While constituting a new criterion for anticlustering, it is surprisingly shown that k-plus anticlustering can be implemented by optimizing the classical k-means criterion after the input data has been augmented with additional variables. A computer simulation and practical examples show that k-plus anticlustering achieves high between-group similarity with regard to multiple objectives. In particular, optimizing between-group similarity with regard to variances usually does not compromise similarity with regard to means; the k-plus extension is therefore generally preferred over classical k-means anticlustering. Examples are given on how k-plus anticlustering can be applied to real norming data using the open source R package `anticlust`, which is freely available via CRAN (https://cran.r-project.org/package=anticlust) and Github (https://github.com/m-Py/anticlust).

@@

# Compute features for k-plus criterion
squared_from_mean <- function(data) {
  apply(data, 2, function(x) (x - mean(x))^2)
}

@@

Anticlustering methods are characterized by (a) an objective function that quantifies between-group similarity and/or within-group heterogeneity, and (b) an algorithm that determines how elements are assigned to groups to maximize the objective function. Several anticlustering objective functions use pairwise dissimilarity ratings such as the Euclidean distance as input. The most prominent criterion of this kind is the *diversity*, which is the total sum of pairwise dissimilarities between elements within the same group [@brusco2019; @gallego2013]. By considering pairwise dissimilarities between elements in the same group, the diversity criterion reflects the total degree of within-group heterogeneity. High within-group diversity simultaneously ensures high between-group similarity.[^equalsizedgroups] Another important anticlustering criterion based on the information in a dissimilarity matrix is the *dispersion*, which is the minimum dissimilarity between any two elements within the same group [@fernandez2013]. Maximizing the dispersion increases within-group heterogeneity by ensuring that any two elements in the same group are as dissimilar from each other as possible. @brusco2019 convincingly argued that anticlustering applications striving for within-group heterogeneity should incorporate both dispersion and diversity, and they presented a bicriterion algorithm to approximate the Pareto efficient set of solutions according to both criteria.

@@

Oftentimes, partitioning applications in psychology are focused on between-group similarity rather than within-group heterogeneity, even though both goals oftentimes coincide. High between-group similarity is for example desirable when designing experimental conditions using different stimulus sets that should be as similar as possible with regard to response-relevant attributes [@lahl2006]. Such stimulus selection tasks are usually conducted on the basis of attribute data and not on pairwise (dis)similarity ratings. That is, individual stimuli are numerically coded on relevant dimensions [@dry2009]. For stimuli in psycholinguistic experiments, attributes may consist of ratings for imagery and concreteness, as well as orthographic variables [@friendly1982]. In other cases, attributes are binary and may represent the presence or absence of a feature [@tversky1977]. When attribute values are available, anticlustering approaches that do not require a dissimilarity matrix can be used to partition the data. K-means is probably the best-known clustering objective that can directly be computed on attribute values. In k-means clustering, the sum of the squared Euclidean distances between data points and their cluster centers is minimized, usually using the k-means heuristic [@jain2010; @steinley2006; @brusco2006branch]. Minimizing the k-means criterion simultaneously maximizes between-cluster separation and within-cluster homogeneity [@aloise2009np]. Reversing the k-means objective function---using maximization instead of minimization---has been recognized as a useful approach when aiming for high between-group similarity [@spath1986; @valev1998; @valev1983]. Specifically, as @spath1986 noted, k-means anticlustering minimizes differences with regard to the means of the numeric attributes across clusters. However, other distribution characteristics---such as the variance---are not targeted. @papenberg2020 showed that k-means anticlustering tends to over-optimize between-group similarity with regard to the mean at the cost of similarity in variance. However, when aiming for overall between-group similarity, neglecting all distribution characteristics other than the mean is misguided; similar means can be obtained even when the underlying distributions are clearly different [e.g., @anscombe1973].

@@

If the $\mathit{SSE}$ is maximal---which is the goal of k-means *anticlustering*---, the cluster centroids are as close as possible to the overall centroid $\overline{\mathbf{x}}$ and therefore to each other [@spath1986]. Thus, k-means anticlustering directly optimizes the similarity of the mean attribute values between clusters.

@@

To obtain an adaption of the $\mathit{SSE}$ that quantifies the degree to which the variance differs between groups, a new variable is computed for each original variable; the new variable represents the squared distance of each of the original values to the overall mean of the variable. We obtain a new data matrix $\mathbf{Y}_{N \times P} = \{y_{ij}\}_{N \times P}$ where $y_{ij} = (x_{ij} - \overline{\mathbf{x}}_j)^2$. On the basis of the data matrix $\mathbf{Y}_{N \times P}$ we can define the criterion $\mathit{SSE_{Var}}$ reflecting how strongly the variances of all attributes vary across the $K$ clusters:

@@

The weights $w_1$ and $w_2$ can be chosen to adjust the relative importance of equalizing means and variances, respectively. Some simplifications can be arranged when ignoring the weights $w_1$ and $w_2$, e.g., by setting them both to 1. This simplification leads to the *unweighted sum approach* for k-plus anticlustering: When defining $z_{i.} = (x_{i1}, \ldots, x_{iP}, y_{i1}, \ldots, y_{iP})$, the bicriterion $\mathit{SSE_{kplus}}$ can be computed as

@@

This formula again has the same form as $\mathit{SSE}$ but is computed on different data. In this case, both the original data $x_{ij}$ as well as the squared distances $y_{ij}$ are used as attributes that enter the objective function. That is, computing the k-plus criterion can be done by first computing the values $y_{ij}$, appending them to the original data matrix, and then computing $\mathit{SSE}$ on the augmented data. Hence, using the unweighted sum approach, k-plus anticlustering leaves the original k-means criterion unchanged. Note that a problem may arise if the weights $w_1$ and $w_2$ are ignored when optimizing the k-plus objective. Because the computation of the new attribute values $y_{ij}$ involves squaring, their range of values is expected to be very different from the raw values $x_{ij}$ (e.g., see Figure 1). This imbalance can strongly and unpredictably influence the relative weight of $\mathit{SSE}$ and $\mathit{SSE_{Var}}$ when computing $\mathit{SSE_{kplus}}$. To ensure an equal weight of both criteria, all attributes ($x_{ij}$ and $y_{ij}$) can be standardized before computing $\mathit{SSE_{kplus}}$, e.g. via $z$-standardization.

@@

```{r, fig.cap = "The relationship between the objectives $\\mathit{SSE}$ and $\\mathit{SSE_{Var}}$ across all ways to partition $N = 14$ normally distributed random values into $K = 2$ groups. The partition represented by the square maximizes $\\mathit{SSE}$ and thus minimizes the difference in means between groups; the partition represented by the triangular maximizes $\\mathit{SSE_{Var}}$ and thus minimizes the difference in variances between groups; the partition represented by the diamond maximizes $\\mathit{SSE_{kplus}}$ and satisfies both constituting criteria to an almost optimal degree.", fig.height = 5, fig.width = 5}

@@

The three k-plus methods were compared to (i) k-means anticlustering, minimizing differences with regard to group means; (ii) diversity anticlustering, maximizing the sum of pairwise Euclidean distances within groups; (iii) a simple random allocation of elements to groups. All anticlustering methods used a $z$-standardization of the data sets as a preprocessing step.

@@

apa_table(
  tab,
  caption = "Results of the simulation study",
  note = "The global results of the simulation study aggregated across all conditions. Cells contain information about the average global between-group discrepancy with regard to means, standard deviations, skewness, kurtosis, and correlations. Lower values indicate less discrepancy, i.e., higher between-group similarity. Each cell value is the result of averaging across 30,000 data points (10,000 data sets $\\times$ $K = 2$, $K = 3$ or $K = 4$).",
  escape = FALSE,
  align = c("r")
)

@@

```{r Figure1, fig.width = 8, fig.height = 10, fig.cap = "Results of the simulation study: Depicts the performance of the anticlustering methods with regard to minimizing discrepancy in means, variances, skewness, kurtosis and correlations (split by $N$, averaged across the remaining variables that varied in the simulation)."}

@@

ldf %>% 
  group_by(method, Objective, N) %>% 
  summarise(Mean = mean(value)) %>% 
  filter(method != "random") %>% 
  mutate(
    Objective = ordered(
      Objective, 
      levels = c("means", "sd", "skew", "kur", "cor"),
      labels = paste0(
        "Delta[italic(",
        c("M", "SD", "Skew", "Kurtosis", "Cor"),
        ")]"
      )
    ) 
  ) %>%
  ggplot(aes(x = N, y = Mean, colour = method)) + 
  geom_line(aes(linetype = method), size = .85) + 
  facet_grid(rows = vars(Objective), scales = "free", labeller = label_parsed) + 
  ylab("Mean discrepancy")+
  xlab(expression(italic(N))) +
  theme_bw(base_size = 16) + 
  scale_x_continuous(
    breaks = axis_ticks, 
    labels = axis_labels
  )

@@

I used the function `kplus_anticlustering()` to divide the data set of 900 images into 9 groups of 100 images each. The first argument of the function call is a variable named `features`, which contains the image data as a table: 900 rows representing images and 3 columns representing the features beauty, arousal, and valence. The accompanying OSF repository contains the full code to reproduce the example (**TODO**).

@@

kplus_anticlustering(
  features, # data table with 900 rows and 3 features
  K = 9,
  variance = TRUE,
  skew = TRUE,
  kurtosis = FALSE,
  covariances = FALSE,
  moments = NULL,
  method = "local-maximum",
  standardize = TRUE
)
```

The function call takes about 10 seconds on a contemporary personal computer; the run time may vary depending on how many iterations are needed to find the local maximum. The function `kplus_anticlustering()` has four boolean arguments to specify whether variance, skewness, kurtosis and covariances should be included as part of the k-plus criterion. Only the argument `variance` is set to `TRUE` by default---corresponding to the understanding that k-plus is an extension to k-means that at least equalizes variances in addition to means. The arguments corresponding to skewness, kurtosis and convariance have to be spefically "turned on", i.e., by setting them to `TRUE` in the function call of `kplus_anticlustering()`. If other higher order moments should be included as part of the optimization, the optional argument `moments` can be used to specify the desired moments as an integer vector. The argument `method = "local-maximum"` ensures that the k-plus criterion is optimized using the local maximum search method described in the previous section. The argument `K` describes the number of groups. By default, the function ensures that all features are standardized before the optimization starts; in particular, this option enforces that all k-plus criteria receive the same weight during the optimization process. This behaviour can be adjusted using the boolean argument `standardize`. The documentation of the `anticlust` package also provides detailed information on how to use the function `kplus_anticlustering()`.

@@

Table 4 illustrates the results that are obtained when using standard[^standardkplus] k-plus anticlustering to divide the stimulus set into 6 groups of size `r nrow(schaper2019) / 6` each. Even though the means and standard deviations are no longer perfectly matched between groups, arguably they are similar enough for any practical purposes.

[^standardkplus]: I refer to *standard* k-plus anticlustering as the minimization of differences in means and variances, but not in higher order moments or covariances.

@@

Anticlustering is a partitioning method that aims for high between-group similarity and high within-group heterogeneity. In this paper, I presented the k-plus criterion for anticlustering, which focuses on maximizing between-group similarity. K-plus represents between-group similarity as discrepancies in distribution moments between groups: means, variances, skewness, kurtosis, higher order moments, and covariances. Depending on the application, a different subset of these objectives can be included in the anticlustering process. The k-plus criterion is an extension of the classical k-means criterion, which only reflects how similar groups are with regard to their means. Interestingly, however, the k-plus extension does not change the original k-means criterion because additional criteria can be incorporated by augmenting the input data.

A simulation study and examples on real norming data showed that the k-plus criterion is well suited to maximize between-group similarity with regard to multiple criteria. Given that similarity with regard to variances was achieved with almost no sacrifice to similarity in means, k-plus anticlustering should be considered as the default replacement of classical k-means when striving for between-group similarity.[^binarydata] The results did however show that fulfilling multiple criteria at the same time can be aggravated in small data sets. In practice, users can try out different objectives and see which specification fulfils their needs. A reasonable approach is to start "greedy", e.g. by specifying that multiple distribution moments should equalized. If the results of such a greedy setup are subpar, users can reduce their requirements by opting for fewer criteria (e.g., only the mean and variance). Using the open source R package `anticlust`, the (repeated) application of k-plus anticlustering is easily accessible and free.
